What's in a Name? \\ Linear Temporal Logic \\ Literally Represents Time Lines

### 2023-08-27 paper/main.tex
fine-toothed typo corrections: pages 1--2. Moving on to page 3 now...

\usepackage[strings]{underscore} % fixes doi underscores
\usepackage{hyperref} % brings up so many errors
\usepackage[capitalise, noabbrev]{cleveref}

@@

@@

We make the following contributions in our paper:
\begin{itemize}
    \item We provide the tool, \tool, which takes \ltl\ formulas and synthesizes timeline visualizations for them.
    \item We prove correctness in \cref{sec:analysis}. We show that for every \ltl\ formula, the timeline outputted indeed represents the set of satisfying traces.
    \item We showcase the results of our tool on a range of example formulas in \cref{sec:tool-showcase} and \cref{sec:playing}, including discussing the air traffic control example above.
\end{itemize}

@@

We begin by setting up the prerequisite definitions involved in our work.
First, we provide the definition and semantics for \ltl. In the following subsections, we define (state-based) \Buchi automata and $\omega$-regular expressions, which are generalizations of the well-known deterministic finite automata (DFAs) and regular expressions, for the case of infinite words. Lastly, we describe the graphic visualizations we will use to depict timelines.

@@

\begin{definition}[Linear Temporal Logic (\ltl)]
    % cite Kristin's survey paper
    The syntax of an \ltl\ formula over a set of atomic propositions $\AP$, where $p\in\AP$ is a propositional variable, is given by the following grammar:
    \begin{align*}
        \varphi ::= & p \mid \neg \varphi \mid \varphi \land \varphi \mid \varphi \lor \varphi \mid \varphi \limplies \varphi \mid \always \varphi \mid \eventually \varphi \mid \nextt \varphi \\ & \mid \varphi \stronguntil \varphi \mid \varphi \weakrelease \varphi
    \end{align*}\label{ltl-defn}
\end{definition}

\begin{remark}
    Intuitively, $\always \varphi$ says that formula $\varphi$ is true at every time step; $\eventually \varphi$ says that formula $\varphi$ is true either now or at sometime in the future time step; and $\nextt \varphi$ says formula $\varphi$ is true at the next time step immediately after the current one; $\varphi \stronguntil \Psi$ says that formula $\varphi$ is true \textit{until} such time formula $\Psi$ becomes true; $\varphi \weakrelease \Psi$ says that formula $\Psi$ must be true now and remain true until formula $\varphi$ becomes true, after which point \textit{release} $\Psi$ (if formula $\varphi$ is never true, then formula $\Psi$ must remain true at all time).
\end{remark}

\begin{remark}
    The tool we provide in this paper uses a set of concrete syntax to represent those logical connectives, in order to avoid the difficulty of typesetting connectives such as $\always$ and $\eventually$. The concrete syntax is defined in~\ref{sec:concrete-syntax}.
\end{remark}

\begin{definition}[Semantics of \ltl\ formula]
    The semantics of \ltl\ formula is defined via  $\pi, i \models \varphi$ where $\pi : \omega \rightarrow 2^{\AP}$ is the computation that stores the truthhood and falsehood of every atomic proposition at every time step, and $\omega$ is the set of natural numbers that denote the time step. The semantics follows exactly from Rozier~\cite{Roz11}.
\end{definition}

\subsection{State-based \Buchi automata (BA)}
\begin{definition}[$\omega$-word]
    An $\omega$-word or infinite run of $\Sigma$, is an infinite string $s = (s_0, s_1, s_2, \dots)$ where each $s_i\in \Sigma$
\end{definition}
\begin{definition}[\Buchi automaton]
    A \Buchi automaton is a $5$-tuple, $G = (Q, \Sigma, \delta, s, F)$ consisting of
    \begin{enumerate}
        \item a finite set of states $Q$
        \item a finite alphabet of input symbols called the alphabet $\Sigma$
        \item a relation $\delta \subseteq (Q\times \Sigma) \times Q$, called the transition relation
        \item an initial or start state called $s\in Q$
        \item a set of accepting states $F \subseteq Q$
    \end{enumerate}
    $G$ accepts an infinite run iff at least one of its infinitely visited states is in $F$
\end{definition}

\subsection{$\omega$-regular expression}
\begin{definition}[Regular expression]\label{def:regex}
    For symbol $a \in \Sigma$, regular expression is defined by the following grammar:
    \begin{align*}
        A ::= \emptyset \mid \epsilon \mid a \mid AA \mid A + A \mid A^*
    \end{align*}
\end{definition}
\begin{definition}[Semantics of regular expression]
    Let $L(A)$ denote the set of propositional formulas accepted by regular expression $A$. Then $L(A)$ is inductively defined as
    \begin{align*}
        L(\emptyset) & = \emptyset \\
        L(\epsilon) & = \{\epsilon\} \tag{$\epsilon$ denotes the empty string}\\
        L(a) & = \{a\} \\
        L(A_1A_2) & = \{s_1s_2 \mid s_1 \in L(A_1) \text{ and } s_2 \in L(A_2)\} \\
        L(A_1 + A_2) & = L(A_1) \cup L(A_2) \\
        L^{(0)}(A) & = \{\epsilon\} \\
        L^{(i + 1)}(A) & = \{s_1s_2 \mid s_1 \in L(A) \text{ and } s_2 \in L^{(i)}(A)\} \\\
        L(A^*) & = \bigcup_{i \ge 0} L^{(i)}(A)
    \end{align*}
\end{definition}

\begin{remark}
    For the purpose of our tool, $\Sigma$ is the set of propositional logic formulas defined as
    \[
    \Psi ::= p \mid \top \mid \bot \mid \neg \Psi \mid \Psi \land \Psi \mid \Psi \lor \Psi \mid \Psi \limplies \Psi
    \]
\end{remark}

@@

\subsection{Timeline}\label{sec:timeline}
We present timelines as graphic visualizations. It contains the following features:
\begin{itemize}
    \item Every timeline starts with a node named ``start''.
    \item Every node represents one time step, and each node has a propositional logic formula $\Psi$, which specifies the behavior of atomic propositions at that time step. The formula $\Psi$ must be true at that time step. If $\Psi = 1$, that means all atomic propositions can behave arbitrarily.
    \item A node in egg shape with a caption ``repeats $0$ - $\infty$'' means to repeat the (one time step) current node for arbitrarily finitely many times.
    \item A node with label ``$\cdots$'' means to repeat the pattern prior to it and after it arbitrarily finitely many times.
    \item The grey box means repeat infinitely. Once we reach the end of a timeline in the grey box, we must reenter the same grey box from any of its (left) starting points.
    \item Timelines can be in parallel, signifying that either of the parallel timeline could happen.
\end{itemize}

@@

\appendix
\section{Concrete Syntax of \tool} \label{sec:concrete-syntax}
Recall from~\ref{ltl-defn} that we define an \ltl\ formula as
\begin{align*}
    \varphi ::= & p \mid \neg \varphi \mid \varphi \land \varphi \mid \varphi \lor \varphi \mid \varphi \limplies \varphi \mid \always \varphi \mid \eventually \varphi \mid \nextt \varphi \\ & \mid \varphi \stronguntil \varphi \mid \varphi \weakrelease \varphi
\end{align*}
where $p \in \AP$. In our tool \tool, we use the following concrete syntax to represent \ltl\ formulas.
\[
    \begin{array}{c r l l l@{\qquad} l}
         &      & \textit{abstract syntax}              & \textit{concrete syntax}                       & \textit{description} \\
    \varphi & ::=  & p                                     & \texttt{p}                                     & \text{atomic prop.} \\
         & \mid & \neg \varphi                             & \texttt{!} \varphi                                & \text{negation} \\
         & \mid & \varphi \land \varphi                       & \varphi \texttt{ \& } \varphi                        & \text{conjunction} \\
         & \mid & \varphi \lor \varphi                        & \varphi \texttt{ | } \varphi                         & \text{disjunction} \\
         & \mid & \varphi \limplies \varphi                   & \varphi \texttt{ -> } \varphi                        & \text{implication} \\
         & \mid & \always \varphi                          & \texttt{G} \varphi                                & \text{globally} \\
         & \mid & \eventually \varphi                      & \texttt{F} \varphi                                & \text{finally} \\
         & \mid & \nextt \varphi                           & \texttt{X} \varphi                                & \text{next} \\
         & \mid & \varphi \stronguntil \varphi                & \varphi \texttt{ U } \varphi                         & \text{until} \\
         & \mid & \varphi \weakrelease \varphi                & \varphi \texttt{ R } \varphi                         & \text{release} \\
    \end{array}
\]
%\nocite{*}
\bibliographystyle{IEEEtran}
\bibliography{RelatedWork}

>>>>

\usepackage[strings]{underscore} % fixes doi underscores
\usepackage{hyperref} % brings up so many errors
\usepackage[capitalise, noabbrev]{cleveref}
\usepackage{mathrsfs} %for mathscr

@@

\renewcommand{\phi}{\varphi}

\renewcommand{\L}{\mathscr{L}}

@@

We make the following contributions in our paper:
\begin{itemize}
    \item We provide the tool, \tool, which takes \ltl\ formulas and synthesizes timeline visualizations for them.
    \item We prove correctness in \cref{sec:analysis}. We show that for every \ltl\ formula, the timeline outputted indeed represents the set of satisfying traces.
    \item We showcase the results of our tool on a range of example formulas from real-life industrial verification efforts in \cref{sec:tool-showcase} and \cref{sec:playing}, including discussing the air traffic control example above.
\end{itemize}

@@

We begin by setting up the prerequisite definitions involved in our work.
First, we provide the definition and semantics for \ltl. In the following subsections, we define (state-based) \Buchi automata and $\omega$-regular expressions, which are generalizations of the well-known deterministic finite automata (DFA) and regular expressions, for the case of infinite words. Lastly, we describe the graphic visualizations we will use to depict timelines.

@@

\begin{definition}[Linear Temporal Logic (\ltl)]
    % cite Kristin's survey paper
    The syntax of an \ltl\ formula over a set of atomic propositions $\AP$, where $p\in\AP$ is a propositional variable, consists of the following grammar:
    \begin{align*}
        \varphi ::= & p \mid \neg \varphi \mid \varphi \land \psi \mid \varphi \lor \psi \mid \varphi \limplies \psi \mid \always \varphi \mid \eventually \varphi \mid \nextt \varphi \\ & \mid \varphi \stronguntil \psi \mid \varphi \weakrelease \psi
    \end{align*}\label{ltl-defn}
\end{definition}

\begin{remark}
    Intuitively, $\always \varphi$ says that formula $\varphi$ is true at every time step; $\eventually \varphi$ says that formula $\varphi$ is true either now or at some time in the future; $\nextt \varphi$ says formula $\varphi$ is true at the next time step immediately after the current one; $\varphi \stronguntil \psi$ says that formula $\varphi$ is true \textit{until} (strictly before) such time formula $\psi$ becomes true (and $\psi$ is either true now or in the future); and $\varphi \weakrelease \psi$ says that formula $\psi$ must be true now and remain true unless formula $\varphi$ becomes true, after which point we can \textit{release} $\psi$ (if formula $\phi$ is never true, then formula $\psi$ must remain true at all times).
\end{remark}

\begin{remark}
    The \tool tool uses a concrete input syntax of ASCII symbols to represent those logical connectives, in order to simplify typesetting connectives such as $\always$ and $\eventually$. For reference, the concrete syntax appears in Appendix ~\ref{sec:concrete-syntax}.
\end{remark}

\begin{definition}[Semantics of \ltl\ ]
    Let $\pi : \omega \rightarrow 2^{\AP}$ be a \emph{computation} or trace that stores the truthhood and falsehood of every atomic proposition at every time step, where $\omega$ is the set of natural numbers that label the time step.
    Then we say $\pi, i \models \varphi$, that is, computation $\pi$ starting from time $i \in \omega$ models LTL formula $\phi$ whenever the result of this evaluation satisfies $\phi$; see Rozier~\cite{Roz11} for the full LTL semantics. 
\end{definition}

\subsection{State-based \Buchi automata (BA)}
\begin{definition}[$\omega$-word]
    Let $\Sigma$ be an alphabet. An $\omega$-word or infinite run of $\Sigma$, is an infinite string $s = (s_0, s_1, s_2, \dots)$ where each $s_i\in \Sigma$.
\end{definition}
\begin{definition}[\Buchi automaton]
    A \Buchi automaton is a $5$-tuple, $B = (Q, \Sigma, \delta, s, F)$ consisting of
    \begin{enumerate}
        \item a finite set of states $Q$
        \item a finite alphabet of input symbols $\Sigma$
        \item a transition relation $\delta \subseteq (Q\times \Sigma) \times Q$
        \item an initial or start state called $s\in Q$
        \item a set of accepting states $F \subseteq Q$
    \end{enumerate}
    $B$ accepts an infinite run iff at least one of its infinitely visited states is in $F$.
\end{definition}

\subsection{$\omega$-regular expressions}
\begin{definition}[Regular expression]\label{def:regex}
    Let $a$ be a symbol in $\Sigma$, $A$ be a regular expression, and $\epsilon$ be the empty expression. We define regular expressions using the following grammar:
    \begin{align*}
        A ::= \emptyset \mid \epsilon \mid a \mid AA \mid A + A \mid A^*
    \end{align*}
\end{definition}
\begin{definition}[Semantics of regular expressions]
    Let $\L(A)$ denote the set of propositional formulas accepted by regular expression $A$, also called the language accepted by $A$. We define $\L(A)$, inductively as:
    \begin{align*}
        \L(\emptyset) & = \emptyset \\
        \L(\epsilon) & = \{\epsilon\} \tag{$\epsilon$ denotes the empty string}\\
        \L(a) & = \{a\} \\
        \L(A_1A_2) & = \{s_1s_2 \mid s_1 \in L(A_1) \text{ and } s_2 \in L(A_2)\} \\
        \L(A_1 + A_2) & = L(A_1) \cup L(A_2) \\
        \L^{(0)}(A) & = \{\epsilon\} \\
        \L^{(i + 1)}(A) & = \{s_1s_2 \mid s_1 \in L(A) \text{ and } s_2 \in L^{(i)}(A)\} \\\
        \L(A^*) & = \bigcup_{i \ge 0} L^{(i)}(A)
    \end{align*}
\end{definition}

\begin{remark}
    For the purpose of our tool, we define $\Sigma$ as the set of propositional logic formulas:
    \[
    a \in \Sigma ::= p \mid \top \mid \bot \mid \neg \phi \mid \phi \land \psi \mid \phi \lor \psi \mid \phi \limplies \psi
    \]
\end{remark}

@@

\subsection{Timeline}\label{sec:timeline}
We present timelines as graphic visualizations. It contains the following features:
\begin{itemize}
    \item Every timeline starts with a node named ``start''.
    \item Every node represents one time step, and each node has a propositional logic formula $\psi$, which specifies the behavior of atomic propositions at that time step. The formula $\psi$ must be true at that time step. If $\psi = 1$, that means all atomic propositions can behave arbitrarily.
    \item A node in egg shape with a caption ``repeats $0$ - $\infty$'' means to repeat the (one time step) current node for arbitrarily finitely many times.
    \item A node with label ``$\cdots$'' means to repeat the pattern prior to it and after it arbitrarily finitely many times.
    \item The grey box means repeat infinitely. Once we reach the end of a timeline in the grey box, we must reenter the same grey box from any of its (left) starting points.
    \item Timelines can be in parallel, signifying that either of the parallel timeline could happen.
\end{itemize}

@@

\appendix
\section{Concrete Syntax of \tool} \label{sec:concrete-syntax}
Recall from~\ref{ltl-defn} that we define an \ltl\ formula as
\begin{align*}
    \varphi ::= & p \mid \neg \varphi \mid \varphi \land \psi \mid \varphi \lor \psi \mid \varphi \limplies \psi \mid \always \varphi \mid \eventually \varphi \mid \nextt \varphi \\ & \mid \varphi \stronguntil \psi \mid \varphi \weakrelease \psi
\end{align*}
where $p \in \AP$ and $\varphi$ and $\psi$ are LTL formulas. In our tool \tool, we use the following concrete input syntax to represent \ltl\ formulas in ASCII text.
\[
    \begin{array}{c r l l l@{\qquad} l}
         &      & \textit{abstract syntax}              & \textit{concrete syntax}                       & \textit{description} \\
    \varphi & ::=  & p                                     & \texttt{p}                                     & \text{atomic prop.} \\
         & \mid & \neg \varphi                             & \texttt{!} \varphi                                & \text{negation} \\
         & \mid & \varphi \land \psi                       & \varphi \texttt{ \& } \psi                        & \text{conjunction} \\
         & \mid & \varphi \lor \psi                        & \varphi \texttt{ | } \psi                         & \text{disjunction} \\
         & \mid & \varphi \limplies \psi                   & \varphi \texttt{ -> } \psi                        & \text{implication} \\
         & \mid & \always \varphi                          & \texttt{G} \varphi                                & \text{globally} \\
         & \mid & \eventually \varphi                      & \texttt{F} \varphi                                & \text{in the future} \\
         & \mid & \nextt \varphi                           & \texttt{X} \varphi                                & \text{next} \\
         & \mid & \varphi \stronguntil \psi                & \varphi \texttt{ U } \psi                         & \text{until} \\
         & \mid & \varphi \weakrelease \psi                & \varphi \texttt{ R } \psi                         & \text{release} \\
    \end{array}
\]
%\nocite{*}
\bibliographystyle{IEEEtran}
\bibliography{RelatedWork}
